A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-08 paper.tex
Pass through gARG section

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a stochastic
process, the classical definition of an ARG data structure is
in terms of \emph{events}. Nodes represent
common ancestor or recombination events that occured in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In order to distinguish this form of ARG from what we will be discussing
in later sections we call it an ``event ARG``, or eARG for brevity.
An example of this form of an eARG
is given in Fig.~\ref{fig-arg-data-structure}B.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage (e.g., node \noderef{e}
in Fig.~\ref{fig-arg-data-structure}), and in a recombination
event a lineage is split into two independent
ancestral lineages (e.g., node \noderef{d}
in Fig.~\ref{fig-arg-data-structure}). The final vital
detail---which distinguishes an arbitrary graph with the required topological
properties from an ARG---is that we associate a breakpoint with each recombination
event. This information is sufficient to uniquely
define the local genealogical trees at every position along the genome,
which is the basic requirement for a data structure encoding
genome-wide genetic ancestry.

@@

\section*{Genome ARGs}
% We're not the first to talk about ARGs and pedigrees, and we're
% not doing anything controversial.
The deep connection between ARGs and pedigrees is
well understood
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}
and recent discussions have illustrated the
relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
showing how the coalescence and recombination events in an ARG
correspond to particular genomes in the pedigree.
As discussed in the previous section, the definition of an ARG
as a data structure in terms of these two \emph{events} is problematic,
fundamentally limiting the patterns of ancestry that can be described.
Following \cite{mathieson2020ancestry}, instead of focussing on events,
we define the ARG data structure in terms of \emph{genomes} and the
patterns of transmission between them.
% Bit of duplication here, but good to make the point about being
% motivated by real existing datasets not some academic desire to
% model complexity that isn't there.
This, coupled with
the explicit association of genome coordinates with edges, allows
us to represent even the bewilderingly complex patterns of relatedness
present in modern datasets.

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An Ancestral Recombination Graph defined in terms of genomes embedded
in a pedigree. (A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal (left) and maternal (right) genomes
labelled \textsf{a} to \textsf{p}. Black lines show transmission paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Two recombination breakpoints, at positions 2 and 7 mean that genomes \textsf{a}
and \textsf{c} are independent recombinants of the paternal genomes \textsf{e}
% Gregor: there is some overlap between a and c still (the middle region inherited
% from f) so, a and c are not fully independent!
and \textsf{f}: in these cases the edge annotations have been shown, indicating the
region of genome transmitted. Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker; this shows that all
genomic regions fully coalesce in genome \textsf{n} in the oldest generation.
(B) The equivalent genome ARG, where there is a single type of node (representing
a genome) along with annotations on all edges; annotations showing partial
transmission are highlighted in bold. Recombination, such as that causing the
breakpoint at position 7, is associated with two parent genomes (\textsf{e} and
\textsf{f}) and a child genome (the recombinant, \textsf{c}).
(C) The edge-annotated genome ARG in (B) fully defines the
local relationships along the genome, as shown in the local
(``marginal'') trees.
}
\end{figure}

To distinguish from the classical event-based encodings,
we call this encoding of an ARG data
structure a ``Genome ARG'', or gARG.
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG.
Fig.~\ref{fig-arg-in-pedigree}A shows the genomes
embedded in an inbred diploid pedigree, in which individuals occur at
discrete generations and carry two genomes (maternal and paternal).
The pedigree is shown here for concreteness, and
there is no inherent limitation on mating system, ploidy, or age structure
in the gARG encoding.
A genome is inherited from one of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination. Fig.~\ref{fig-arg-in-pedigree}B shows the resulting gARG,
depicting the ancestry of the sample genomes \textsf{a}--\textsf{d}. The graph
topology along with the genome coordinate annotations on the
edges then uniquely defines the trees at every position
along the genome (Fig.~\ref{fig-arg-in-pedigree}C).

@@

The focus on genomes rather than events has some interesting
and important ramifications. Firstly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{h} is the direct ancestor of \textsf{d} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{d} descends directly from \textsf{k} (see
the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
a representational perspective. This ability can be useful in practice and could be
% Do we need to spell this out? Gregor: I think so - see below
required if, for example, we have a deeply sequenced pedigree as in \cite{RosFreixedes2022}.
% e.g. Honeybees? Be great to cite an example!
% Note by Yan - if we actually have the sequence, then this turns into
% a sample node, so would be normally included anyway. Perhaps, therefore
% we should talk about these ``pass though'' nodes in the context
% of internal samples.
% Gregor: There are no good deeply sequenced pedigrees in honeybees, but sex
% chromosome behaves like haplodiploid system, also any deep pedigree works here, so I cited our pig
% work. Also, I think its important to not think of all genomes as current samples - we have many
% generations of animals with genome or SNP array data and we know pedigree so we can't assume they
% are all current samples - I believe you made this clear distinction also in the tsdate paper with
% ancient human samples

Secondly, unlike in an eARG, nodes in a gARG are not classified into different
types. Indeed a single node can have both multiple children
and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
for instance, node \textsf{a} were to have two children.
% It is a shame that this is not actually shown in fig 3, but I don't think
% we can do this unless we have 5 generations
Moreover, a single node can have more than 2 children, and (as we shall see in
the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
state this is that transitions between nodes in a gARG can encapsulate multiple
events in an eARG.

The final point deals with the way in which recombination is represented, and is
closely linked to which genomes are chosen as the nodes in a gARG.
We could potentially choose genomes at any point in the life cycle: for instance,
we could chose to represent gametes as ARG nodes. However, the most intuitive
choice is where genomes represent diploid individuals in each generation, as in
Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
is not represented by a single ``recombination node'' (as in an eARG), but with
two parent genomes between which recombination takes place
(e.g.\ nodes \textsf{e} and \textsf{f}), and a
single resulting ``recombinant'' genome (e.g.\ \textsf{c}). The
edges that link this recombinant with its two parents carry the information
concerning breakpoint location that would be associated with a
recombination node in an eARG. For example, node \textsf{c} in
Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{f}
and positions 7 to 10 from \textsf{e}, indicating that the breakpoint is at position 7.

For simplicity, throughout the rest of the paper we will contrast the
classical eARG (as defined in the XXX section) with the gARG definition
here. However, it is important to note that many of the benefits of the
gARG encoding that we discuss are due to the encoding of inheritance via
edges rather than nodes (see section XXX).
Whether one regards nodes as genomes or events
is something of a philosophical question, and mostly unimportant
from a computational perspective.

@@

@@

\begin{itemize}
\item It is unfortunate that we are adding yet another meaning to the
term ARG here, but all we're really doing is making the implicit
assumptions that others have made (i.e., describing the output of
Relate, tsinfer etc as ARGs) and making them concrete. We need to have a
term to describe all of these things, and since people are already using
ARG for this, we might as well make it rigorous. We can then move away
from the unfruitful discussion about which methods infer a ``true'' or
``real'' ARG (none do, really, under the original definitions) and
instead start discussing the \emph{properties} of these ARGs. (See
next point also.)
\item Terminology based around an inferred ancestry being a ``true''
ARG is impoverished and misleading. The only reasonable interpretation
of a ARG being true or not is whether it's a sample from the coalescent.
Most inference methods (and all that scale well) do not. In any case,
the coalescent is just one model, and a highly idealised one. It is
exceedingly unlikely that the actual ARG describing the ancestry
of any biological population is ``true'' ARG in this sense. The terminology is
impoverished because it reduces the possibilities available
to something either being an ARG or not, whereas there are many different
interesting properties of these structures that we should be able to
discuss and compare.
\item tskit can represent any ARG.
\item Interchange is a real problem, and if we're going to realise the
potential for ARGs in genomics we must solve it. Efforts to standardise
phylogenetic networks haven't done very well (~\citep{cardona2008extended}
has 89 citations). The only real effort to standardise an ARG format
based on extending GraphML~\citep{mcgill2013graphml} has 11 citations.
GraphML~\citep{mcgill2013graphml} has some of the same key ideas
as tskit though, in particular allowing the annotation of ``live sites''
to an edge as a set of disjoint intervals that hold ancestral material.

>>>>

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a stochastic
process, the classical definition of an ARG data structure is
in terms of \emph{events}. Nodes represent
common ancestor or recombination events that occured in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In order to distinguish this form of ARG from what we will be discussing
in later sections we refer to it as an ``event ARG``, or eARG for brevity.
An example of this form of an eARG
is given in Fig.~\ref{fig-arg-data-structure}B.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage (e.g., node \noderef{e}
in Fig.~\ref{fig-arg-data-structure}), and in a recombination
event a lineage is split into two independent
ancestral lineages (e.g., node \noderef{d}
in Fig.~\ref{fig-arg-data-structure}). The final vital
detail---which distinguishes an arbitrary graph with the required topological
properties from an ARG---is that we associate a breakpoint with each recombination
event. This information is sufficient to uniquely
define the local genealogical trees at every position along the genome,
which is the basic requirement for a data structure encoding
genome-wide genetic ancestry.

@@

\section*{Genome ARGs}
As we have seen in previous sections, the focus
on a limited palette of events makes describing currently-available
datasets and known patterns of genetic inheritance as an eARG
quite cumbersome.
There is no fundamental reason that the patterns of genetic inheritance
that we wish to describe via an ARG must be in terms
of discrete \emph{events}, however. Recent discussions have illustrated the
relationship between ARGs and
the pedigree~\citep{mathieson2020ancestry,brandt2021evaluation},
and the deep connections between the structures are
well understood
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}.
Following \cite{mathieson2020ancestry},
rather defining the inheritance
of genetic material in terms of \emph{events} as we do in an eARG,
we can instead define an ARG data structure in terms of
\emph{genomes} and intervals of genetic inheritance
along the edges (see the previous section).
To distinguish from the classical Griffiths eARG,
we refer to this encoding of an ARG data structure
as a ``Genome ARG'', or gARG.
The distinction between nodes representing events
and genomes may seem pedantic at first, but as we
see in later sections, we gain a great deal of flexibility
and computational efficiency by decoupling our representation
of genetic ancestry from the events that generated it.

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An Ancestral Recombination Graph defined in terms of genomes embedded
in a pedigree. (A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal (left) and maternal (right) genomes
labelled \textsf{a} to \textsf{p}. Black lines show transmission paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Two recombination breakpoints, at positions 2 and 7 mean that genomes \textsf{a}
and \textsf{c} are independent recombinants of the paternal genomes \textsf{e}
% Gregor: there is some overlap between a and c still (the middle region inherited
% from f) so, a and c are not fully independent!
and \textsf{f}: in these cases the edge annotations have been shown, indicating the
region of genome transmitted. Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker; this shows that all
genomic regions find a common ancestor in genome \textsf{n} in the oldest generation.
(B) The corresponding genome ARG, where there is a single type of node (representing
a genome) along with annotations on all edges; annotations showing partial
transmission are highlighted in bold. Recombination (e.g.,
the cause of the breakpoint at position 7)
is associated with two parent genomes (\textsf{e} and
\textsf{f}) and a child genome (the recombinant, \textsf{c}).
(C) The edge-annotated genome ARG in (B) fully defines the
local relationships along the genome, as shown in the local
(``marginal'') trees.
}
\end{figure}

A gARG is a graph in which nodes represent monoploid genomes (i.e.,
haploids have one genome, diploids two, etc.) and edges represent
genetic inheritance between an ancestor and a descendant.
Each edge is annotated (see the previous section) with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance).
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
showing how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
A diploid pedigree is shown here for concreteness, but
there is no limitation on mating system, ploidy, or age structure
in the gARG encoding.
In the diploid setting, a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty.

@@

There are some interesting and important consequences to this focus on genomes
rather than events.
Firstly, a gARG may contain nodes that are both samples \emph{and}
internal nodes in the local trees.
This ability is useful when we have pedigree information
along with genetic data
\cite[e.g.][]{hayes20191000,RosFreixedes2022,anderson2022genes},
where we may have many generations of internal sample nodes
whose genomes have been sequenced.
Another situation in which it is important that
we do not assume samples are all contemporary ``leaf'' nodes
(or that all internal nodes are either coalescences or recombinations)
is when we are incorporating ancient genomes into ARG
inference~\citep{speidel2021inferring,wohns2022unified}.
Similarly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{h} is the direct ancestor of \textsf{d} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{d} descends directly from \textsf{k} (see
the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
a representational perspective.

Secondly, because we do not classify nodes by event type
(common ancestor or recombination) there is no limitation
on the complexity what can be modelled---multiple recombinations
and coalescences can happen simultaneously in the same genome.
In large deeply sequenced pedigrees such complexities are commonplace.

Finally, there is a subtle and important point about how recombination
is represented: which genome in the organismal lifecycle
do nodes represent?
In principle, we can intepret nodes as representing genomes
at any point in the life cycle [supplementary Figure ?]. Hudson
views nodes as representing gametes~\citep{hudson1983properties}.
However, we argue that the most intuitive
approach is to let nodes represent the genomes present in diploid
individuals in each generation, as in
Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
is not represented by a single ``recombination node'' (as in an eARG), but with
two parent genomes between which recombination takes place
(e.g.\ nodes \textsf{e} and \textsf{f}), and a
single resulting ``recombinant'' genome (e.g.\ \textsf{c}). The
edges that link this recombinant with its two parents carry the information
concerning breakpoint location that would be associated with a
recombination node in an eARG. [TODO finish up here with more
clarity on what this means in terms of the required specifity
of recombination events, and put in a forward ref to the simplification
section where we talk about stacking up recombination events
and their identifiability.]

For simplicity, throughout the rest of the paper we will contrast the
classical eARG (as defined in the XXX section) with the gARG definition
here. However, it is important to note that many of the benefits of the
gARG encoding that we discuss are due to the encoding of inheritance via
edges rather than nodes (see section XXX).
Whether one regards nodes as genomes or events is orthogonal,
and something of a philosophical question which makes little
difference from a computational perspective.

@@

% JK: Text moved from an earlier section. See if we want to reuse this here
% if not already said.
% Indeed a single node can have both multiple children
% and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
% for instance, node \textsf{a} were to have two children.
% % It is a shame that this is not actually shown in fig 3, but I don't think
% % we can do this unless we have 5 generations
% Moreover, a single node can have more than 2 children, and (as we shall see in
% the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
% state this is that transitions between nodes in a gARG can encapsulate multiple
% events in an eARG.

@@

\begin{itemize}
\item [Recap quickly distinction bewtween aARG and gARG. Then]
    This, coupled with
    the explicit association of genome coordinates with edges, allows
    us to represent even the bewilderingly complex patterns of relatedness
    present in modern datasets.

\item It is unfortunate that we are adding yet another meaning to the
term ARG here, but all we're really doing is making the implicit
assumptions that others have made (i.e., describing the output of
Relate, tsinfer etc as ARGs) and making them concrete. We need to have a
term to describe all of these things, and since people are already using
ARG for this, we might as well make it rigorous. We can then move away
from the unfruitful discussion about which methods infer a ``true'' or
``real'' ARG (none do, really, under the original definitions) and
instead start discussing the \emph{properties} of these ARGs. (See
next point also.)
\item Terminology based around an inferred ancestry being a ``true''
ARG is impoverished and misleading. The only reasonable interpretation
of a ARG being true or not is whether it's a sample from the coalescent.
Most inference methods (and all that scale well) do not. In any case,
the coalescent is just one model, and a highly idealised one. It is
exceedingly unlikely that the actual ARG describing the ancestry
of any biological population is ``true'' ARG in this sense. The terminology is
impoverished because it reduces the possibilities available
to something either being an ARG or not, whereas there are many different
interesting properties of these structures that we should be able to
discuss and compare.
\item tskit can represent any ARG.
\item Interchange is a real problem, and if we're going to realise the
potential for ARGs in genomics we must solve it. Efforts to standardise
phylogenetic networks haven't done very well (~\citep{cardona2008extended}
has 89 citations). The only real effort to standardise an ARG format
based on extending GraphML~\citep{mcgill2013graphml} has 11 citations.
GraphML~\citep{mcgill2013graphml} has some of the same key ideas
as tskit though, in particular allowing the annotation of ``live sites''
to an edge as a set of disjoint intervals that hold ancestral material.
